### 2018-06-26 latex/introduction.tex

%%% TeX-master: "manuscript"

\section{Introduction}

@@

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD-synthesis pathway.  If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.  In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1).

We here present a phylogenetic analys of NAD-pathway in eukaryotes that. In contrast to our previous analysis, which has been lmited by the low number of eularyotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotich pathway evolution. The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost. This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes. To explain these counterintuitive results we built a mathematical model of the pathway and demonstratet that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam. Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.  Based on multiple sequence alignments,  we identified a sequence insertion in NamPRT in Deuterostomes  that is affecting the affinity of NamPRT,  undermining the predictions derived from our mathematical modelling approach

>>>>

%%% TeX-master: "manuscript"
% !TeX spellcheck=en_GB
\section{Introduction}

@@

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD-synthesis pathway.  If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.  In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1).

We here present a phylogenetic analysis of NAD-pathway in eukaryotes that. In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution. The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost. This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes. To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam. Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.  Based on multiple sequence alignments,  we identified a sequence insertion in NamPRT in Deuterostomes  that is affecting the affinity of NamPRT,  undermining the predictions derived from our mathematical modelling approach

### 2018-06-26 latex/results.tex

%%% TeX-master: "manuscript"

\section{Results}

@@

Looking acroos all known species,   two different pathways exist that recylce NAM. The major pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate. Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.

@@

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1? \todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}. The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa. With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

@@

First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families. We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.  We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B). The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).

These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.  The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concetration.

@@

It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absecne of NNMT. In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell devision rates.

@@

From these observations we derived two possible hypotheses regarding the role of the loop for NamPRT function. Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme. We thus performed immunoflourescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C). We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict. We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E. coli.  Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP. It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.  We furthermore showed  that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (Figure 6E).

@@

The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria? To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material). One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.  Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B). At high cell division rates, steady state concentrations in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems co-exist. 
In the presence of NNMT the NamPRT compartment outcompetes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D). This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations. The NAD consumption flux is still  higher  in the NamPRT and NNMT containing compartment. Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.

>>>>

%%% TeX-master: "manuscript"
\section{Results}

@@

Looking acroos all known species,   two different pathways exist that recylce NAM. The major pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amidation step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate. Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.

@@

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1 \todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}. The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa. With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

@@

First we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families. We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.  We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as expected, reducing the steady state concentration of NAD (Figure 3B). The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).

These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.  The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concentration.

@@

It is interesting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT. In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell division rates.

@@

From these observations we derived two possible hypotheses regarding the role of the loop for NamPRT function. Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme. We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C). We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict. We thus recombinantly expressed  both wildtype and mutant  NamPRT  in E. coli.  Upon purification we measured the activity using NMR detection of NMN produced in the presence and absence of ATP. It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.  We furthermore showed  that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (Figure 6E).

@@

The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria? To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material). One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.  Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B). At high cell division rates, steady state concentrations in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems co-exist. 
In the presence of NNMT the NamPRT compartment out-competes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D). This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations. The NAD consumption flux is still  higher  in the NamPRT and NNMT containing compartment. Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.
